Challenging Fuel Cycle Modeling Assumptions: Facility and Time Step Discretization Effects

### 2016-02-05 paper.tex
i.e./e.g. -> i.e.,/e.g.,

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different because of inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e. individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics will
allow for making more deliberate decisions with respect to trade-offs such as
computational investment vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
\emph{stocks} representing isotopes or isotope groups that have quantities that change
over time because of \emph{flows} between them.  The flows are determined by equations
that are a function of the state of the system (i.e. the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics based simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

System dynamics based fuel cycle simulators generally use a collection of
stocks to represent the state of groups of like-facilities commonly referred
to as "fleets".  As the simulation steps through time, the levels of stocks
(e.g. reactor fresh fuel inventory, repository waste inventories, etc.) are
adjusted according to the calculated flow values.  Fuel cycle simulators that model
facility fleets built on a system dynamics foundation include VISION
\cite{jacobson_verifiable_2010}, DANESS \cite{van_den_durpel_daness_2009},
DYMOND, and ORION \cite{gregg_orion_2012} among
others.  In fleet-based modeling:

@@

%% response to technical comment 1.
These four cases span a spectrum of fidelity with one end being more realistic
(i.e. smaller time steps, individual facilities) and the other being more
appropriate for scoping studies (i.e.  larger time steps, fleet facilities).
Conducting this analysis involved developing some additional capability
in/around Cyclus.  Three primary pieces to the methodology are:

@@

A fleet reactor archetype was created for Cyclus.  Although the modeling of
individual, discrete facilities was a motivation for Cyclus, its flexibility
makes it relatively straightforward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to continuous
flow fleet facilities.  In order to contrast well with the individual,
single-facility granularity modeling that is part of the standard Cyclus
archetypes, a fleet archetype was created for this work with aggregate
facility behavior and continuous flow.  Although it would certainly be
interesting to compare many intermediate levels of facility discretization
(e.g. 1-reactor groups, 10-reactor groups, single-fleet, etc.), such a
detailed analysis is beyond the scope of this work.
% address technical comment 14

@@

    \item Reactor capacity is deployed and retired in increments of single
      reactor units, each capable of producing power $P_r$ (e.g. 1 GWe
      LWRs).  $N_{r}$ reactors are deployed at any time capable of producing
      up to $P_r \cdot N_r$ power.

@@

The described scenario is used as closely as possible in each of the four
cases with appropriate adjustments for case-specific differences (e.g. monthly
vs. quarterly time steps).  Invariants preserved with respect to reactor
behavior between all four cases are shown in Table \ref{tab:invar}.  Table
\ref{tab:reactor-detail} shows the computed/selected configuration for both
LWR and SFR reactor types that was used in Cyclus input files for all four
cases.

@@

In each of the four cases, the initial 100 LWRs are modified to have a
zero-length refueling outage with an explicit 900 MWe net power output
capacity.  Because the individual reactor model does not have any way to start
a reactor mid-cycle, this is a modeling trick that was done to avoid having
all initial LWRs refuel at the same time.  This adjustment serves to improve
the realism associated with the effects being investigated (i.e. fleet-average
power generation) even though it is unrealistic in other ways.
% response to technical comment 11

@@

% response to technical comment 12
Instead of annual deployments, a longer 21-month deployment period is used
because it provides better natural staggering for the reactor refueling
cycles. The 630-month lowest common multiple (LCM) for 21-months (deployment
period), 15 months (SFR cycle), and 18 months (LWR cycle) is much larger than
the 180-month LCM when using a 12-month build period providing a more even
power generation profile.  The same deployment schedule is used for all four
cases; power capacity is built along a curve starting at 100 GWe in year zero
following the 1\% exponential growth curve.  Although many alternate
deployment schedules exist that avoid SFR fuel shortages (e.g. a more gradual
shift toward SFRs), this particular schedule is deliberately chosen because it
causes significant fuel shortages.  This is done in order to force the
simulation and facility models into more extreme circumstances in order to
expose behavioral differences among the design choices investigated.
% response to general comment paragraph
% response to technical comment 13

@@

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot P_r$
of power for that time step where $P_r$ is the power capacity of a single
reactor of the fleet (e.g.~1000~MWe).  If an individually modeled reactor is
short one half batch of fuel, then it will produce no power ($0 \cdot P_r$) that
time step.  Time step duration can amplify this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
Consider an example of a 3-batch reactor with a fuel shortage of $\frac{1}{2}$ batch of
fuel that would be resolved in 1 month.  Depending on the modeling choice, each reactor
would produce the following amounts of energy over a 3 month period:

@@

Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e. cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
partly be a result of quantized shutdown.

@@

Another difference between fleet-based and individual facility modeling
involves fuel sharing.  Power production by a group of individual reactors for
a fixed amount of fuel shortage may vary depending on how available fuel is
distributed.  This is illustrated in Figure \ref{fig:fuel-sharing}.  A system
being short 3 batches might mean that one reactor has no batches and is not
operating or that 3 reactors are each missing a single batch.  This is
referred to as the \emph{fuel sharing effect}.  Fleet reactors, by design,
have perfect fuel sharing --- all fuel is distributed so that the minimum
possible amount of capacity is offline during shortages.  The most correct
measure of fuel sharing inefficiency is the difference between the actual
generated power and the maximum amount of power that could have been generated
among all possible fuel distribution alternatives.  This is difficult to
actually compute. One possible approximation is to count the number of fresh
fuel batches distributed to reactors that ended up not having enough fuel to
operate (i.e. a full core) that time step.  This approximation assumes that
every "wasted" batch of fuel, could have been given to a reactor that only
needed one batch to operate.  Equation \ref{eq:badshare} describes such an
approximation.

@@

One important phenomenon related to time step duration is the \emph{inventory
drawdown effect}.  Increasing time step duration reduces the frequency over
which refueling can occur, resulting in larger impulse drawdown on
inventories. This by itself is not problematic because this is balanced by
correspondingly larger inventory top-up quantities.  However, another
component of this effect is the simultaneity of incoming and outgoing material
flows on a given time step.  Within a particular time step facilities do not
know about incoming material when they make commitments for outgoing material.
For longer time steps, larger incoming quanta of material are not available
for making offers. In general longer time steps create a need for larger
floating inventory buffers in order to avoid material shortages.  Another
aspect of this effect is that at least one time step is required for a
material object to move between two facilities. Larger time steps result in
longer minimum bounds on time to traverse paths between facilities. Although
direct measures of this effect are difficult, some of its consequences (e.g.
higher standing inventories during shortages) can be quantified by comparing
scenarios with different time step durations (e.g.  cases MF and QF).

There are also performance differences between the two modeling paradigms.  A
single 200 year simulation with several hundred fleet-based reactors takes
about 2 seconds to run on a computer with an Intel Core(TM) i7-4770 3.40GHz CPU where the same simulation
using individual reactor facilities takes about 20 seconds.  This order of
magnitude difference is an important trade-off to keep in mind when doing fuel
cycle analysis in general, particularly when considering optimization or
sensitivity studies where running thousands or millions of simulations may be
desirable.  The time step duration has less impact on performance because the
amount of work being done in the simulation is similar; in the monthly time
step cases most facilities (i.e. all reactors) only perform actions every
refueling cycle rather than every time step.

@@

One useful overestimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by the fast reactor power capacity
(i.e. 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative energy deficit caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases MI and QI respectively.  These
approximations account for a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepancy in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

@@

    \item Multiple fuel quanta comprise a single batch (i.e. multiple
        assemblies per batch). This functionality is natively supported in the
        individual reactor model.

@@

Also notable is that different modeling choices (e.g. time step duration,
facility discretization, etc.) may have varying levels of significance
depending on the scenario/context they are operating in.  As shown in Figure
\ref{fig:puinv-compare} before the fuel shortage begins, differences between
each of the four cases are relatively minor.  However, after the shortage
begins near year 100, separated Pu inventories vary much more significantly
between the different cases.

>>>>

The diversity of assumptions embedded within various fuel cycle simulators pose challenges for direct
comparison and benchmarking.  In 2012 the Organisation for
Economic Co-operation and Development (OECD) completed a benchmark study
\cite{oecd2012benchmark} that is perhaps one of the most complete published
comparisons performed.  Despite this, however, various results from the
different simulators were often significantly different because of inconsistent
modeling decisions involving reprocessing strategies, refueling behavior,
reactor end-of-life handling, etc.  This work attempts to identify and
quantify the effects of selected modeling choices relative to facility modeling
discretization (i.e., individual facilities vs. aggregate fleets) and time step
duration.  A better understanding of how these affect fuel cycle dynamics will
allow for making more deliberate decisions with respect to trade-offs such as
computational investment vs. realism.

@@

Traditionally, many fuel cycle simulators have used a system dynamics approach
\cite{forrester_industrial_1961} to modeling. These simulators model several
\emph{stocks} representing isotopes or isotope groups that have quantities that change
over time because of \emph{flows} between them.  The flows are determined by equations
that are a function of the state of the system (i.e., the system's stocks).  A
constructed system is then solved in discrete or continuous time depending on
the limitations of the software and constraints imposed by the model.  In
system dynamics based simulators, the number of stocks and potential flows between
them is generally a static property of the modeled system, making it somewhat
difficult to swap different reactor models in and out of the same scenario
without wiring them into the simulator manually.

System dynamics based fuel cycle simulators generally use a collection of
stocks to represent the state of groups of like-facilities commonly referred
to as "fleets".  As the simulation steps through time, the levels of stocks
(e.g., reactor fresh fuel inventory, repository waste inventories, etc.) are
adjusted according to the calculated flow values.  Fuel cycle simulators that model
facility fleets built on a system dynamics foundation include VISION
\cite{jacobson_verifiable_2010}, DANESS \cite{van_den_durpel_daness_2009},
DYMOND, and ORION \cite{gregg_orion_2012} among
others.  In fleet-based modeling:

@@

%% response to technical comment 1.
These four cases span a spectrum of fidelity with one end being more realistic
(i.e., smaller time steps, individual facilities) and the other being more
appropriate for scoping studies (i.e.,  larger time steps, fleet facilities).
Conducting this analysis involved developing some additional capability
in/around Cyclus.  Three primary pieces to the methodology are:

@@

A fleet reactor archetype was created for Cyclus.  Although the modeling of
individual, discrete facilities was a motivation for Cyclus, its flexibility
makes it relatively straightforward to create plug-in archetypes to model
anywhere along the spectrum from individual, discrete facilities to continuous
flow fleet facilities.  In order to contrast well with the individual,
single-facility granularity modeling that is part of the standard Cyclus
archetypes, a fleet archetype was created for this work with aggregate
facility behavior and continuous flow.  Although it would certainly be
interesting to compare many intermediate levels of facility discretization
(e.g., 1-reactor groups, 10-reactor groups, single-fleet, etc.), such a
detailed analysis is beyond the scope of this work.
% address technical comment 14

@@

    \item Reactor capacity is deployed and retired in increments of single
      reactor units, each capable of producing power $P_r$ (e.g., 1 GWe
      LWRs).  $N_{r}$ reactors are deployed at any time capable of producing
      up to $P_r \cdot N_r$ power.

@@

The described scenario is used as closely as possible in each of the four
cases with appropriate adjustments for case-specific differences (e.g., monthly
vs. quarterly time steps).  Invariants preserved with respect to reactor
behavior between all four cases are shown in Table \ref{tab:invar}.  Table
\ref{tab:reactor-detail} shows the computed/selected configuration for both
LWR and SFR reactor types that was used in Cyclus input files for all four
cases.

@@

In each of the four cases, the initial 100 LWRs are modified to have a
zero-length refueling outage with an explicit 900 MWe net power output
capacity.  Because the individual reactor model does not have any way to start
a reactor mid-cycle, this is a modeling trick that was done to avoid having
all initial LWRs refuel at the same time.  This adjustment serves to improve
the realism associated with the effects being investigated (i.e., fleet-average
power generation) even though it is unrealistic in other ways.
% response to technical comment 11

@@

% response to technical comment 12
Instead of annual deployments, a longer 21-month deployment period is used
because it provides better natural staggering for the reactor refueling
cycles. The 630-month lowest common multiple (LCM) for 21-months (deployment
period), 15 months (SFR cycle), and 18 months (LWR cycle) is much larger than
the 180-month LCM when using a 12-month build period providing a more even
power generation profile.  The same deployment schedule is used for all four
cases; power capacity is built along a curve starting at 100 GWe in year zero
following the 1\% exponential growth curve.  Although many alternate
deployment schedules exist that avoid SFR fuel shortages (e.g., a more gradual
shift toward SFRs), this particular schedule is deliberately chosen because it
causes significant fuel shortages.  This is done in order to force the
simulation and facility models into more extreme circumstances in order to
expose behavioral differences among the design choices investigated.
% response to general comment paragraph
% response to technical comment 13

@@

One difference between the fleet and individual facility modeling is the
\emph{quantized shutdown effect}.  If a fleet-based reactor modeling 3-batch
cores is short one half batch of fuel, it will produce $\frac{2.5}{3}\cdot P_r$
of power for that time step where $P_r$ is the power capacity of a single
reactor of the fleet (e.g.,~1000~MWe).  If an individually modeled reactor is
short one half batch of fuel, then it will produce no power ($0 \cdot P_r$) that
time step.  Time step duration can amplify this effect.  For both fleet and
individual reactor modeling, power goes offline in full time step increments.
Consider an example of a 3-batch reactor with a fuel shortage of $\frac{1}{2}$ batch of
fuel that would be resolved in 1 month.  Depending on the modeling choice, each reactor
would produce the following amounts of energy over a 3 month period:

@@

Quantifying this effect directly is difficult.  One way of observing
it is to compare fleet reactor scenarios that have different time step
durations (i.e., cases MF and QF).  Because the fleet reactors have perfect fuel
sharing (described below) and no noise from refueling outages, differences observed will
partly be a result of quantized shutdown.

@@

Another difference between fleet-based and individual facility modeling
involves fuel sharing.  Power production by a group of individual reactors for
a fixed amount of fuel shortage may vary depending on how available fuel is
distributed.  This is illustrated in Figure \ref{fig:fuel-sharing}.  A system
being short 3 batches might mean that one reactor has no batches and is not
operating or that 3 reactors are each missing a single batch.  This is
referred to as the \emph{fuel sharing effect}.  Fleet reactors, by design,
have perfect fuel sharing --- all fuel is distributed so that the minimum
possible amount of capacity is offline during shortages.  The most correct
measure of fuel sharing inefficiency is the difference between the actual
generated power and the maximum amount of power that could have been generated
among all possible fuel distribution alternatives.  This is difficult to
actually compute. One possible approximation is to count the number of fresh
fuel batches distributed to reactors that ended up not having enough fuel to
operate (i.e., a full core) that time step.  This approximation assumes that
every "wasted" batch of fuel, could have been given to a reactor that only
needed one batch to operate.  Equation \ref{eq:badshare} describes such an
approximation.

@@

One important phenomenon related to time step duration is the \emph{inventory
drawdown effect}.  Increasing time step duration reduces the frequency over
which refueling can occur, resulting in larger impulse drawdown on
inventories. This by itself is not problematic because this is balanced by
correspondingly larger inventory top-up quantities.  However, another
component of this effect is the simultaneity of incoming and outgoing material
flows on a given time step.  Within a particular time step facilities do not
know about incoming material when they make commitments for outgoing material.
For longer time steps, larger incoming quanta of material are not available
for making offers. In general longer time steps create a need for larger
floating inventory buffers in order to avoid material shortages.  Another
aspect of this effect is that at least one time step is required for a
material object to move between two facilities. Larger time steps result in
longer minimum bounds on time to traverse paths between facilities. Although
direct measures of this effect are difficult, some of its consequences (e.g.,
higher standing inventories during shortages) can be quantified by comparing
scenarios with different time step durations (e.g.,  cases MF and QF).

There are also performance differences between the two modeling paradigms.  A
single 200 year simulation with several hundred fleet-based reactors takes
about 2 seconds to run on a computer with an Intel Core(TM) i7-4770 3.40GHz CPU where the same simulation
using individual reactor facilities takes about 20 seconds.  This order of
magnitude difference is an important trade-off to keep in mind when doing fuel
cycle analysis in general, particularly when considering optimization or
sensitivity studies where running thousands or millions of simulations may be
desirable.  The time step duration has less impact on performance because the
amount of work being done in the simulation is similar; in the monthly time
step cases most facilities (i.e., all reactors) only perform actions every
refueling cycle rather than every time step.

@@

One useful overestimate of the fuel sharing inefficiency is to assume every
one of the "wasted" batches could have been given to a reactor enabling it to
operate.  Multiplying each of these cumulative batches by the fast reactor power capacity
(i.e., 450 MWe) provides a way to compare fuel sharing inefficiency with the
overall fuel outage.  This results in a cumulative energy deficit caused by
poor fuel sharing of roughly 2,000 $\text{GWe} \cdot \text{months}$ and 6,600
$\text{GWe} \cdot \text{months}$ for cases MI and QI respectively.  These
approximations account for a significant portion of the difference between individual
and fleet cases in Figure \ref{fig:unfueled} suggesting that most of the
discrepancy in cumulative outage between fleet and individual modeling is
caused by inefficient fuel sharing in the individual reactor cases.

@@

    \item Multiple fuel quanta comprise a single batch (i.e., multiple
        assemblies per batch). This functionality is natively supported in the
        individual reactor model.

@@

Also notable is that different modeling choices (e.g., time step duration,
facility discretization, etc.) may have varying levels of significance
depending on the scenario/context they are operating in.  As shown in Figure
\ref{fig:puinv-compare} before the fuel shortage begins, differences between
each of the four cases are relatively minor.  However, after the shortage
begins near year 100, separated Pu inventories vary much more significantly
between the different cases.
